Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-04-25 paper/main.tex
some citations
changes introduction

\abstract{\textbf{Motivation:} Current state of the art gene annotation programs for eukaryotic genomes have to be pretrained for each species and still often produce annotations that lack the required accuracy. \\
\textbf{Results:} We set up a framework that can be used to build much more general gene callers that also outperform traditional methods on a variety of complex genomes.\\
\textbf{Availability:} Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text Text  Text\\
\textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

\section{Introduction}
For many tasks, it is necessary to have DNA regions labeled according to their function. This process is called gene annotation and it can be performed in different levels of precision, from simple coding -- non-coding classification to detailed structural labeling, such as enhancer regions or long non-coding RNA sequences in eukaryotes. In this work on eucaryotic genomes, we differentiate between Intergenic regions, Untranslated regions (UTRS, parts at the beginning and end of gene, that are transcribed but not translated), Coding sequences (CDS) and Introns. Because of the sheer size of genomes alone, gene annotation is a very labour intensive process and genomic database providers like NCBI \citep{thibaud2013eukaryotic} and Ensembl \citep{aken2016ensembl} offer gene annotation pipelines. These pipelines integrate experimental data (from e.g. RNA-seq or proteogenomics) with homolog sequences in the database and ab-initio gene predictions. The latter is an attractive approach, because it is cheap and fast. State of the art performance ab-initio gene prediction is achieved by higher order hidden markov models (HMMs), such as Genscan \citep{burge1997prediction} or Augustus \citep{stanke2003gene} but their accuracy leaves room for improvement. By choosing possible states and transitions, designers of HMMs assume structure in the sequence that may limit its predictive power. Indeed, in practice HMMs have problems with generalizing across species and it seems actual learning of underlying patterns is limited. In the last decade, deep neural networks have been applied with tremendous success in many areas of statistical modelling, including biology \citep{ching2018opportunities}. For sequence data, such as DNA, recurrent neural networks (RNN) in the form of long-short term memory (LSTM) are an established architecture and here, we employ them to improve on HMMs for ab-initio genome annotation.

RNNs have already shown promising results in related, but generally more limited investigations. \citep{choudharypredicting} carries out preliminary explorations on the potential of BLSTMs for cross species gene prediction, but trains his model on human genes only and tests it on two more species. DeepAnnotator \citep{amin2018deepannotator} uses bidirectional LSTMs (BLSTM) for gene finding in prokaryotes. Gene prediction in prokaryotes is considered more amenable than in eukaryotes, as genes in prokaryotes are proportionately more frequent in the genome, feature simpler control structures and do not use splicing \citep{wang2004brief}. DanQ \citep{quang2016danq} proposes the use of a BLSTM after a concurrent neural net (CNN) to find detailed motifs in the human genome. DeePromotor \citep{oubounyt2019deepromoter} trains a similar architecture for the recognition of promotor regions. Recently, several groups \citep{jaganathan2019predicting, wang2019splicefinder} successfully used CNNs to find splicing sites.

\section{Approach}
Helixer is a prototype software for training and utilizing a general purpose model for the functional cross-species annotation of large eukaryotic genomes using only DNA data as input. We demonstrate the effectiveness of this approach by training a single model for the annotation of a large set of genomes from the kingdoms metazoa and viridiplantae, respectively, which we will call {\it animals} and {\it plants} from now on. In total we used 192 animal genomes and 60 plant genomes. In contrast to traditional approaches, we employ modelling techniques from the field of deep learning to improve both the annotation accuracy and the ability to generalize across genomes. The source code and all data is publically available.

\begin{methods}
\section{Methods}
We worked with 192 animal and 60 plant genomes. The data of each genome consists of the latest publically available genomic assembly in form of a FASTA file and the latest annotation in the GFF format. We used all animals genomes in EnsemblMetazoa 45 \citep{howe2020ensembl} as well as all non-embargoed plant genomes of the JGI Phytosome 13 database \citep{goodstein2012phytozome}. Both data groups were used seperately during training as well as during evaluation. At the start of our work we split both genome groups into {\it training genomes} and {\it test genomes}. The test genome group is made up of 19 and 6 genomes, respectively, and was chosen to be of the best possible quality based on collected metadata while also representing a broad phylogenetic spread. We then experimented which genomes to use for during training from the pool of non-test genomes while evaluating the performance of models on the remaining ones. We call the genomes that were ultimatily not used during training the {\it validation genomes}. While we did split off a group of test genomes at the beginning of our work, we ended up only reporting on the generalization performance of the combined set of validation and test genomes as the validation genomes are by far the largest set. We also did not find a noticible difference in performance between those two groups.

@@

This work has been supported by the... Text Text  Text Text.\vspace*{-12pt}

>>>>

\abstract{\textbf{Motivation:} Current state of the art gene annotation programs for eukaryotic genomes have to be pretrained for each species and still often produce annotations that lack the required accuracy. \\
\textbf{Results:} We set up a framework that can be used to build much more general gene callers that also outperform traditional methods on a variety of complex genomes.\\
\textbf{Availability:} \\
\textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

\section{Introduction}
The annotation of genes in DNA is fundamental for many other tasks depending on it to be done as good and correct as possible. It can be performed at different levels of precision, from simple coding -- non-coding classification to detailed structural labeling. Because of the sheer size of genomes alone, gene annotation is a very labour intensive process and genomic database providers like NCBI \citep{thibaud2013eukaryotic} and Ensembl \citep{aken2016ensembl} offer gene annotation pipelines. These pipelines integrate experimental data (from e.g. RNA-seq or proteogenomics) with homolog sequences in the database and ab-initio gene predictions. The latter is an attractive approach, because it is cheap and fast. State of the art performance is achieved by higher order hidden markov models (HMMs), such as Genscan \citep{burge1997prediction} or AUGUSTUS \citep{stanke2003gene} but their accuracy leaves room for improvement. By encoding possible states and transitions in a probalistic model, designers of HMMs assume structure in the sequence that may limit its predictive power. Indeed, in practice HMMs have problems with generalizing across species and it seems actual learning of underlying patterns is limited despite the fundamental ability to do so being present.

In the last decade, deep neural networks (DNN) have been applied with great success in many areas of statistical modelling, including biology \citep{ching2018opportunities}. For sequence data, such as DNA, speech or text, a special kind of recurrent neural network (RNN) called long-short term memory (LSTM) \citep{hochreiter1997long} is an established building block for many different architectures. LSTM units can also be used to process sequential input starting from both ends, forming a bi-directional LSTM (BLSTM). It has also been shown that HMMs can be successfully combined with DNNs \citep{liu2016novo, liu2016pedla}.

For the purpose of gene annotation, RNNs have already shown promising results. \citep{choudharypredicting} carries out preliminary explorations on the potential of BLSTMs for cross species gene prediction and trains his model on human genes to test it later on two more species. DeepAnnotator \citep{amin2018deepannotator} uses bidirectional LSTMs (BLSTM) for gene finding in prokaryotes. Gene prediction in prokaryotes is considered more amenable than in eukaryotes, as genes in prokaryotes are proportionately more frequent in the genome, feature simpler control structures and do not use splicing \citep{wang2004brief}. DanQ \citep{quang2016danq} proposes the use of a BLSTM after a concurrent neural net (CNN) to find detailed motifs in the human genome. DeePromotor \citep{oubounyt2019deepromoter} trains a similar architecture for the recognition of promotor regions. Recently, several groups \citep{jaganathan2019predicting, wang2019splicefinder} successfully used CNNs to find splicing sites.

In this work we present Helixer, a novel prototype software for training and utilizing a general purpose DNN for the ab-initio cross-species gene annotation of large eukaryotic genomes using only DNA data as input. Our model is trained to differentiate between 4 regions: Intergenic, Untranslated (UTR), Coding (CDS) and Intron. We demonstrate the effectiveness of this approach by training a single model for the annotation of a large set of genomes from the kingdoms metazoa and viridiplantae, respectively, which we will call {\it animals} and {\it plants} from now on. We worked with the full data of 192 animal genomes and 60 plant genomes. Both the ability to generalize across genomes as well as the scope of the evaluation greatly surpasses all other work in this field. The source code and all input data is publically available.

\begin{methods}
\section{Methods}
The foundation of our work are 192 animal and 60 plant genomes. The data of each genome consists of the latest publically available genomic assembly in form of a FASTA file and the latest annotation in the GFF format. We used all animals genomes in EnsemblMetazoa 45 \citep{howe2020ensembl} as well as all non-embargoed plant genomes of the JGI Phytosome 13 database \citep{goodstein2012phytozome}. Both data groups were used seperately during training as well as during evaluation. At the start of our work we split both genome groups into {\it training genomes} and {\it test genomes}. The test genome group is made up of 19 and 6 genomes, respectively, and was chosen to be of the best possible quality based on collected metadata while also representing a broad phylogenetic spread. We then experimented which genomes to use for during training from the pool of non-test genomes while evaluating the performance of models on the remaining ones. We call the genomes that were ultimatily not used during training the {\it validation genomes}. While we did split off a group of test genomes at the beginning of our work, we ended up only reporting on the generalization performance of the combined set of validation and test genomes as the validation genomes are by far the largest set. We also did not find a noticible difference in performance between those two groups.

@@

This work has been supported by the... \vspace*{-12pt}
